\section{Adding Support for Thread-Local Variables}
\label{sec:thread-local-vars}

% General information about the GetContext effect and that it's used to implement thread-local variables

So far we have looked at a protocol \(\proto{2}\) that has two effects which suffice to model fibers that can suspend and fork off new fibers.
But fibers in Eio can use an additional effect called \egetctx{} that we discuss in this section.
For each fiber the scheduler keeps track of context metadata, one part of which are \emph{thread-local variables}.
Thread-local variables are state that is shared between all fibers of one scheduler (hence thread-local) and a fiber gets access to them via the \egetctx{} effect.

% Example of how thread-local variables can be used for a tracing log.

Since all fibers of one scheduler execute concurrently on one system-level thread, they have exclusive access to the thread-local variables while they are running.
This allows a practical form of shared state without the overhead of synchronization primitives of multithreaded data structures.
Two example use-cases are per-scheduler tracing of events, where all fibers of one scheduler write to a common log,
and inter-fiber message passing, where fibers use a simple queue to exchange messages.
Of course, this comes with the restriction that it is only usable for fibers running in the same thread.

% What we want to verify about thread-local variables. 

In Eio thread-local variables are represented by a dictionary from variable names to arbitrary values and expose an intended API that only allows adding new entries.
However, it is still possible for fibers to arbitrarily modify the whole dictionary, so for demonstration purposes we model thread-local variables as a single mutable reference that is part of the context record: \ocamlin{ctx.tlv}.
Properties we want to prove about thread-local variables are:
\begin{enumerate}
    \item Each time a fiber performs a \egetctx{} effect it receives the same reference.
    \item As long as a fiber does not perform other effects like \efork{} or \esuspend{}, it holds exclusive ownership of the reference.
\end{enumerate}

Code examples illustrating the properties are shown in figure~\ref{fig:tlv-example}.
Note that these are only the most basic properties showing that \ocamlin{ctx.tlv} acts like a normal reference, but one that can be accessed via an effect.
To enable modular proofs of concrete fibers using thread-local variables, we include in our logical definition a predicate \(T\) on the stored value that can be instantiated by fibers as needed.

\begin{figure}[ht]
    \begin{minted}{ocaml}
let fiber1 () =
  let ctx = perform (GetContext ()) in
  let ctx2 = perform (GetContext ()) in
  assert (ctx.tlv == ctx2.tlv)

let fiber2 () =
  let ctx = perform (GetContext ()) in
  let v = !ctx.tlv in
  (* some computation that does not perform Fork/Suspend *)
  ...
  assert (!ctx.tlv == v)
  \end{minted}
    \caption{Constructed example of safety for thread-local variables.}
    \label{fig:tlv-example}
\end{figure}

\subsection{Changes to Logical State}

% How do we extend the protocol and change the logical definitions.
To handle thread-local variables in our development we must change both the implementation and logical definitions.
The necessary changes to the implementation are trivial, so we just refer to the mechanization\footnote{TODO insert link}.
The new definitions are described in figure~\ref{fig:logical-state-ext}.
\(\gsTLVAg{}\; \delta\; tlv\) is used to show the uniqueness of the location \(tlv\).
\(\gsIsFiberContext{}\; \delta\; tlv\) represents the context that is tracked for each fiber, where \(\delta\) is a shorthand for multiple ghost names.
It expresses that the location \(tlv\) is a thread-local variable which maps to some value \(v\) satisfying \(T\).
The predicate T is hidden behind a \(\gsSavedPred{}\) indirection to make the mechanization easier.
\(\gsFiberResources{}\; \delta\) is then used to abstract away the concrete location \(tlv\).
This predicate represents all resources that a fiber owns while it is running, so each fiber specification now has this as a precondition.
Finally, we must change the definition of \(\gsReady{}\) to require \(\gsFiberResources{}\) as a precondition because it is needed to invoke the continuations saved in the scheduler's run-queue.

\begin{figure}[ht]
    \begin{align*}
        \gsTLVAg{}\; \delta\; tlv \triangleq\;            & \ownGhost{\delta}{agree(tlv)}  \qquad \textrm{Persistent}(\gsTLVAg{\delta}{tlv})                          \\
        \gsIsFiberContext{}\; \delta\; tlv\; \triangleq\; & \gsTLVAg{}\; \delta\; tlv \ast \exists T\; v.\; tlv \mapsto v \ast \gsSavedPred{}\; \delta\; T \ast T\; v \\
        \gsFiberResources{}\; \delta\; \triangleq\;       & \exists\; tlv.\; \gsIsFiberContext{}\; \delta\; tlv\;                                                     \\
        \gsReady{}\; \delta\; f \triangleq\;              & \gsFiberResources{}\; \delta\; \wand{} \ewp{f\; ()}{\bot}{\top}
    \end{align*}
    \caption{Logical state definitions for the verification of our Eio model.}
    \label{fig:logical-state-ext}
\end{figure}

The effect protocols of \efork{} and \esuspend{} are amended so that they pass \(\gsFiberResources{}\) from a fiber to the scheduler and from there to the next running fiber via the protocol pre- and postconditions as shown in figure~\ref{fig:coop-protocol-ext}.
The \efork{} effect now also passes the concrete reference that should be used as the thread-local variable of the new fiber.
A fiber uses the \egetctx{} effect to receive the fiber context value and a copy of \(\gsTLVAg{}\).
This is used to show that the reference \(ctx.tlv\) is equal to the one from \(\gsFiberResources{}\) that the fiber already owns so that the contained points-to predicate can be used.

The crux is that now the protocol \proto{} is parameterized by the ghost name \(\delta\) that identifies the thread-local variable.
This so that both the fiber and the scheduler agree on this ghost name.

\begin{figure}[ht]
    \begin{equation*}
        \proto{}~\delta \triangleq \begin{aligned}[t]
            Fork\;       & \#\; \begin{aligned}[t]
                                     & !\; tlv\; e\; ((tlv, e))\; \begin{aligned}[t]
                                                   & \big\{ \gsFiberResources{}\; \delta\; T \ast \gsTLVAg{}\; \delta\; tlv\; \ast                              \\
                                                   & \later\; (\gsFiberResources{}\; \delta\; T \wand{} \ewp{e}{Coop}{\gsFiberResources{}\; \delta\; T}) \big\}
                                              \end{aligned} . \\
                                     & ?\; ()\; \{ \gsFiberResources{}\; \delta\; T \}
                                \end{aligned}                                 \\
            Suspend\;    & \#\; \begin{aligned}[t]
                                     & !\; reg\; W\; (reg)\; \{\gsFiberResources{}\; \delta\; T \ast \gsIsReg{}\; reg\; W\}. \\
                                     & ?\; y\; (y)\; \{ \gsFiberResources{}\; \delta\; T \ast W\; y \}
                                \end{aligned} \\
            GetContext\; & \#\; !\; ()\; \{\top\}.\; ?\; ctx\; (ctx)\; \{ \gsTLVAg{}\; \delta\; ctx.tlv \}
        \end{aligned}
    \end{equation*}
    \caption{Definition of extended \proto{} protocol with \efork{}, \esuspend{}, and \egetctx{} effects.}
    \label{fig:coop-protocol-ext}
\end{figure}

These changes suffice to prove the safety of the two examples in figure~\ref{fig:tlv-example}.

